\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-15 paper/algorithm.tex
paper: cleaned up histograms in algos
paper: minor revisions to algorithm section

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value in any attribute that they directly operate on.

\subsection{Imputation operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$.
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered.

@@

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.
Figure \ref{fig:query-schematic} shows a simple schematic of what a query involving three tables might look like, absent any imputation operators.

\begin{figure}
\Tree [.$\pi$ [.$GroupBy$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators)}
\label{fig:query-schematic}
\end{figure}

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans using a cost model \[\textsc{Cost}(Q) = (1 - \alpha) \times \textsc{Time}(Q) + \alpha \times \textsc{Loss}(Q).\] $\textsc{Time}(Q)$ is an estimate of the runtime of the query derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast
as possible (at the expense of accuracy).

In order to correctly estimate $\textsc{Time}(Q)$ and $\textsc{Loss}(Q)$, the system must have cardinality estimates for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans

@@

@@

\subsection{Imputation Strategies}

>>>>

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the estimated quality of the results.
This plan is subject to two constraints (Sec.~\ref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as it is treated as a black-box operation.

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$.
The imputation algorithm is discussed in detail in Sec.~\ref{sec:imputation}.
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

\subsection{Search space}
\label{sec:search-space}
To keep the query plan search space tractable, only plans that fit the following template are considered.

@@

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
Figure~\ref{fig:query-schematic} shows the layout of a query involving three tables, absent any imputation operators.

\begin{figure}
\Tree [.$\pi$ [.$g$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators).}
\label{fig:query-schematic}
\end{figure}

\subsection{Cost model}
\label{sec:cost-model}
\todo{Explain that loss is now averaged}
\todo{Do we want to keep the behavior that imputation increases in quality with the number of cells, or switch to number of tuples.}
We rank query plans using a cost model \[\textsc{Cost}(q) = (1 - \alpha) \times \textsc{Time}(q) + \alpha \times \textsc{Loss}(q).\] $\textsc{Time}(q)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates. $\textsc{Loss}(q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance. $\alpha = 1.0$ means that the optimal query should have the highest quality imputation (at the expense of performance), $\alpha=0.0$ means that the optimal query should be as fast as possible (at the expense of imputation quality).

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of $q$ and its sub-queries.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

\begin{algorithm}
\input{algorithms/histogram-helpers.tex}
\end{algorithm}

@@

\subsection{Imputation Strategies}
\label{sec:imputation}
